---
title: "Gene Signature Analysis"
format: html
---

```{r}

library(dplyr)
library(readxl)
library(GSVA)
library(ggplot2)
library(singscore)
library(survival)
library(ggsurvfit)
library(mclust)

ggplot2::theme_set(ggplot2::theme_classic())

```


# Load Data 
```{r}

source("functions/Load_Normalized_Expression.R")
normalized_data = load_normalized_expression("../outputs/normalized_expression.csv")
rm(load_normalized_expression)

```


# Epithelial & Mesenchymal Signatures
```{r}

#Make a signatures list
signatures = list()

#Read in signatures
signatures$patient_EMT = read_excel("../references/Tan_2014_EMT_Signatures.xlsx", sheet = "Tumor")

signatures$cell_line_EMT = read_excel("../references/Tan_2014_EMT_Signatures.xlsx", sheet = "Cell_Line")

#Separate into epithelial and mesenchymal signatures
signatures$patient_epi = signatures$patient_EMT |> 
  dplyr::filter(Category == "Epi") |> 
  pull("Gene_Symbol")

signatures$patient_mes = signatures$patient_EMT |> 
  dplyr::filter(Category == "Mes") |> 
  pull("Gene_Symbol")

signatures$cell_epi = signatures$cell_line_EMT |> 
  dplyr::filter(Category == "Epi") |> 
  pull("Gene_Symbol")

signatures$cell_mes = signatures$cell_line_EMT |> 
  dplyr::filter(Category == "Mes") |> 
  pull("Gene_Symbol")

#Create a signature out of overlaps
signatures$signatures_epi_overlap = signatures$patient_epi[signatures$patient_epi %in% signatures$cell_epi]
  
signatures$signatures_mes_overlap = signatures$patient_mes[signatures$patient_mes %in% signatures$cell_mes]

#Clean up
signatures = signatures[-c(1,2)]
  

```

# Prep data
```{r}

#Change column names of normalized data from ensembl ids to gene symbols
colnames(normalized_data) = mapIds(org.Hs.eg.db,
                                    keys = colnames(normalized_data),
                                    keytype = "ENSEMBL",
                                    column = "SYMBOL")

```


# GSVA
```{r}

#GSVA
gsva_results = gsva(expr = as.matrix(t(normalized_data)),
                    gset.idx.list = signatures)

#Make a dataframe of results
gsva_results = as.data.frame(t(gsva_results))

#Add sample metadata
gsva_results = merge(gsva_results, 
                     sample_metadata, 
                     by.x = "row.names",
                     by.y = "sample_id")


```

## Plots
```{r}

##MAKE INTO A FUNCTION##

ggplot(gsva_results, aes(x = patient_epi, y = patient_mes, color = tumor_descriptor)) +
  geom_point() +
  scale_color_manual(values = c(palette_discrete[1], palette_discrete[5])) +
  labs(x = "Epithelial", y = "Mesenchymal", title = "Patient Tumor Derived Signatures") +
  theme_classic()

ggplot(gsva_results, aes(x = cell_epi, y = cell_mes, color = tumor_descriptor)) +
  geom_point() +
  scale_color_manual(values = c(palette_discrete[1], palette_discrete[5])) +
  labs(x = "Epithelial", y = "Mesenchymal", title = "Cell Line Derived Signatures") +
  theme_classic()

ggplot(gsva_results, aes(x = signatures_epi_overlap, y = signatures_mes_overlap, color = tumor_descriptor)) +
  geom_point() +
  scale_color_manual(values = c(palette_discrete[1], palette_discrete[5])) +
  labs(x = "Epithelial", y = "Mesenchymal", title = "Overlapping Derived Signatures") +
  theme_classic()


```

Plot survival by scores; try a combined score of epi - mes

# ssGSEA
```{r}

#ssGSEA
ssgsea_results = gsva(expr = as.matrix(t(normalized_data)), 
                      gset.idx.list = signatures, 
                      method = 'ssgsea')

#Make a dataframe of results
ssgsea_results = as.data.frame(t(ssgsea_results))

#Add sample metadata
ssgsea_results = merge(ssgsea_results, 
                     sample_metadata, 
                     by.x = "row.names",
                     by.y = "sample_id")


```

## Plots
```{r}

ggplot(ssgsea_results, aes(x = patient_epi, y = patient_mes, color = tumor_descriptor)) +
  geom_point() +
  scale_color_manual(values = c(palette_discrete[1], palette_discrete[5])) +
  labs(x = "Epithelial", y = "Mesenchymal", title = "Patient Tumor Derived Signatures") +
  theme_classic()

ggplot(ssgsea_results, aes(x = cell_epi, y = cell_mes, color = tumor_descriptor)) +
  geom_point() +
  scale_color_manual(values = c(palette_discrete[1], palette_discrete[5])) +
  labs(x = "Epithelial", y = "Mesenchymal", title = "Cell Line Derived Signatures") +
  theme_classic()

ggplot(ssgsea_results, aes(x = signatures_epi_overlap, y = signatures_mes_overlap, color = tumor_descriptor)) +
  geom_point() +
  scale_color_manual(values = c(palette_discrete[1], palette_discrete[5])) +
  labs(x = "Epithelial", y = "Mesenchymal", title = "Overlapping Derived Signatures") +
  theme_classic()

```

## Survival

### Divide by GMM

```{r}

#divide into singscore high vs low using a Guassian Mixture Model
gmm_results = Mclust(ssgsea_results$patient_mes, G = 2)
ssgsea_results$patient_mes_gmm = gmm_results$classification

ggplot(ssgsea_results, aes(x = patient_mes, fill = as.factor(patient_mes_gmm))) +
  geom_density()

ssgsea_results = ssgsea_results |> 
  mutate(patient_mes_gmm_group = case_when(patient_mes_gmm == "1" ~ "Low",
                                           patient_mes_gmm == "2" ~ "High",
                                           .default = "unknown"))


#Survival
ssgsea_results |> 
  dplyr::filter(tissue_rollup == "Cutanous" & tumor_descriptor == "Metastatic") |> 
  survfit2(Surv(last_event, survial_censor) ~ patient_mes_gmm_group , data = _) |>
    ggsurvfit() +
    add_pvalue() +
    add_risktable(risktable_stats = "n.risk")

#Cox PH
ssgsea_results |> 
  dplyr::filter(tissue_rollup == "Cutanous" & tumor_descriptor == "Metastatic") |> 
  coxph(Surv(last_event, survial_censor) ~ patient_mes_gmm_group, data = _)

ssgsea_results |> 
  dplyr::filter(tissue_rollup == "Cutanous" & tumor_descriptor == "Metastatic") |> 
  coxph(Surv(last_event, survial_censor) ~ patient_mes_gmm_group + as.integer(stage) + year_of_diagnosis, data = _)


```

Oddly, mesenchymal score high patients seem to have better overall survival. Is the gene signature not appropriate? Am I missing a confounder? 
Try with epi signatures instead

### Divide at Median
```{r}

#try high and low at median
ssgsea_results = ssgsea_results |> 
  mutate(patient_mes_division = ifelse(patient_mes > median(patient_mes), "High", "Low"))

ggplot(ssgsea_results, aes(x = patient_mes, fill = as.factor(patient_mes_division))) +
  geom_density()

ssgsea_results |> 
  dplyr::filter(tissue_rollup == "Cutanous" & tumor_descriptor == "Metastatic") |> 
  survfit2(Surv(last_event, survial_censor) ~ patient_mes_division , data = _) |>
    ggsurvfit() +
    add_pvalue() +
    add_risktable(risktable_stats = "n.risk")


```

```{r}

summary(lm(patient_mes ~ year_of_diagnosis, data = ssgsea_results))

ggplot(ssgsea_results, aes(x = year_of_diagnosis, y = patient_mes)) +
  geom_point() +
  geom_smooth(method = lm)

coxph(Surv(last_event, survial_censor) ~ year_of_diagnosis, data = ssgsea_results)

ssgsea_results |> 
  dplyr::filter(tissue_rollup == "Cutanous" & tumor_descriptor == "Metastatic") |> 
  ggplot(aes(x = year_of_diagnosis, y = last_event, color = as.factor(survial_censor))) +
    geom_point() +
    scale_color_manual(values = c("grey", "black")) +
    geom_smooth(method = lm, linetype = "dashed", size = 0.5)


```


#### Epithelial
```{r}

ssgsea_results = ssgsea_results |> 
  mutate(patient_epi_division = ifelse(patient_epi > median(patient_epi), "High", "Low"))

ggplot(ssgsea_results, aes(x = patient_epi, fill = as.factor(patient_epi_division))) +
  geom_density()

ssgsea_results |> 
  dplyr::filter(tissue_rollup == "Cutanous" & tumor_descriptor == "Metastatic") |> 
  survfit2(Surv(last_event, survial_censor) ~ patient_epi_division , data = _) |>
    ggsurvfit() +
    add_pvalue() +
    add_risktable(risktable_stats = "n.risk")


```

# Singscore of combined epithelial and Mesenchymal
```{r}

#make an empty data frame for results
#rank data
ranked_df = rankGenes(t(normalized_data))
#run singscore
ss_patient_signature = simpleScore(ranked_df, upSet = signatures$patient_mes, downSet = signatures$patient_epi)
ss_cell_signature = simpleScore(ranked_df, upSet = signatures$cell_mes, downSet = signatures$cell_epi)
ss_cell_signature = simpleScore(ranked_df, upSet = signatures$signatures_mes_overlap, downSet = signatures$signatures_epi_overlap)

#add sample metadata
ss_patient_signature = merge(ss_patient_signature, 
                     sample_metadata, 
                     by.x = "row.names",
                     by.y = "sample_id")


```

## Plots
```{r}

ggplot(ss_patient_signature, aes(x = tumor_descriptor, y = TotalScore, color = tissue_rollup)) +
  geom_jitter(width = 0.25) +
  scale_color_manual(values = palette_discrete) +
  labs(x= "", y = "SingScore", title = "Patient Tumor Derived Signature") +
  theme_classic() 

# add mean bar, p-value, 95% CI, number of samples per group, center title

#Cutaneous sites only
ss_patient_signature |> 
  dplyr::filter(tissue_rollup == "Cutanous") |> 
  ggplot(aes(x = tumor_descriptor, y = TotalScore, color = tumor_descriptor)) +
  geom_jitter(width = 0.2) +
  stat_summary(aes(group = tumor_descriptor), fun.y = mean, geom = 'crossbar', width = 0.3, color = "black") +
  scale_color_manual(values = c(palette_discrete[5], palette_discrete[1])) +
  labs(x= "", y = "SingScore", title = "Patient Tumor Derived Signature - Only Cutaneous Samples", subtitle = "Higher Scores Correspond to More Mesenchymal Like") +
  theme_classic() +
  theme(legend.position = "none")

ss_patient_signature |> 
  dplyr::filter(tissue_rollup == "Cutanous") |> 
  ggplot(aes(x = ajcc_pathologic_stage, y = TotalScore, color = tumor_descriptor)) +
  geom_jitter(width = 0.2) +
  stat_summary(aes(group = tumor_descriptor), fun.y = mean, geom = 'crossbar', width = 0.3, color = "black") +
  scale_color_manual(values = c(palette_discrete[5], palette_discrete[1])) +
  labs(x= "", y = "SingScore", title = "Patient Tumor Derived Signature - Only Cutaneous Samples", subtitle = "Higher Scores Correspond to More Mesenchymal Like") +
  theme_classic() 


```

## Survival
```{r}

#filter to primary cutanoues

#divide into singscore high vs low; try using gmm
gmm_results = Mclust(ss_patient_signature$TotalScore)
ss_patient_signature$gmm = gmm_results$classification

ggplot(ss_patient_signature, aes(x = TotalScore, fill = as.factor(gmm))) +
  geom_density()

ss_patient_signature |> 
  dplyr::filter(tissue_rollup == "Cutanous") |> 
  survfit2(Surv(last_event, survial_censor) ~ gmm , data = _) |>
    ggsurvfit() +
    add_pvalue() +
    add_risktable(risktable_stats = "n.risk")





```

## Is year of diagnosis a confounder?

#To do:
make a list of signature scores and add to that
Average ssGSEA, Singscore and GSVA scores
Visualize tissue rollup (e.g. liver, LN)
Control for stage
Filter to cutaneous sites only
Fix cutaneous spelling
Fix "GESA" to GSEA
Bayessian statistics on singscore values
filter to cutaneous only and do diff exp to derive a metastatic signature

# plot all data for correlations